Climate indicators

### 2023-03-03 DRAFT_climate_indicators.Rmd
climate trend figure

*Author and date:* <!-- Write your name here -->

@@

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
#library(tidyverse)
library(dplyr)
library(ggplot2)
library(stars)
library(sf)
library(tmap)
library(lubridate)
library(ggpubr)
library(microbenchmark)
library(tictoc)
library(ncmeta)
library(parallel)
library(readr)
dir <- substr(getwd(), 1,2)

@@

This chapters describes a workflow for generating or preparing indicators based on interpolated climate data from [SeNorge](https://senorge.no/).

@@

```{r}
# setNams dont work on stars proxy obejct, so I use rename instead
lookup <- setNames("mean", paste0("v_", temp_names))

@@

```{r, eval=F}
names(real_temp_summer)
```

```{r, eval=F}
dim(real_temp_summer)
```

```{r, eval=F}
st_get_dimension_values(real_temp_summer, "Year")
```

@@

@@

@@

```{r, eval=F}
(
regional_means <- rast(y_var_norm[,,]) %>%
  exact_extract(reg, fun = 'mean', append_cols = "region", progress=T) %>%
  setNames(c("region", names(y_var_norm)))
  )
```

@@

@@

@@

```{r}
(
regional_means_long <- regional_means %>%
  as.data.frame() %>%
  rename(referenceUpper = reference_upper,
         referenceLow = reference_low,
         referenceLow2 = reference_low2,
         thresholdLow = threshold_low,
         thresholdHigh = threshold_high) %>%
  pivot_longer(!region) %>%
  separate(name, into=c("type", "year"), remove=F)
)
```

```{r}
temp3 <- temp2 %>%
  st_intersection(reg)
```

>>>>

*Author and date:* 
Anders L. Kolstad
March 2023

@@

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
#library(tidyverse) # cant update to v2 on server
library(dplyr)
library(ggplot2)
library(stars)
library(sf)
library(tmap)
library(lubridate)
library(ggpubr)
library(microbenchmark)
library(tictoc)
library(ncmeta)
library(parallel)
library(readr)
library(exactextractr)
library(tidyr)
library(terra)
dir <- substr(getwd(), 1,2)

@@

This chapters describes the development of a workflow for generating or preparing indicators based on interpolated climate data from [SeNorge](https://senorge.no/).

@@

```{r}
# setNames dont work on stars proxy obejct, so I use rename instead
lookup <- setNames("mean", paste0("v_", temp_names))

@@

@@

```{r, eval=T}
names(real_temp_summer)
```

```{r, eval=T}
dim(real_temp_summer)
```

```{r, eval=T}
st_get_dimension_values(real_temp_summer, "Year")
```

@@

The attribute name is *mean*. We can change this to something more meaningful.

@@

```{r, eval=F}
(
regional_means <- rast(y_var_norm) %>%
  exact_extract(reg, fun = 'mean', append_cols = "region", progress=T) %>%
  setNames(c("region", names(y_var_norm)))
  )
```

@@

I should also get the sd:

```{r, eval=F}
(
regional_means_sd <- rast(y_var_norm) %>%
  exact_extract(reg, fun = function(X) c(mean = mean(x), sd = sd(x)), append_cols = "region", progress=T) %>%
  setNames(c("region", names(y_var_norm)))
  )
```

@@

```{r}
regional_means <- readRDS("temp/regional_means.rds")
```

@@

```{r, warning=F}
div <- c("reference_upper",
         "reference_low",
         "reference_low2",
         "threshold_low",
         "threshold_high",
         "sd")

temp <- regional_means %>%
  as.data.frame() %>%
  select(region, div)

(
regional_means_long <- regional_means %>%
  as.data.frame() %>%
  select(!all_of(div)) %>%
  pivot_longer(!region) %>%
  separate(name, into=c("type", "year")) %>%
  pivot_wider(#id_cols = region,
              names_from = type)  %>%
  left_join(temp, by=join_by(region)) %>% 
  mutate(diff = v-reference_upper) %>%
  mutate(threshold_low_centered = threshold_low-reference_upper) %>%
  mutate(threshold_high_centered = threshold_high-reference_upper)
)

```{r}
regOrder = c("Østlandet","Sørlandet","Vestlandet","Midt-Norge","Nord-Norge")

regional_means_long %>%
  mutate(col = if_else(diff>0, "1", "2")) %>%
  ggplot(aes(x = as.numeric(year), 
           y = diff, fill = col))+
  geom_bar(stat="identity")+
  geom_hline(aes(yintercept = threshold_low_centered),
        linetype=2)+
  geom_hline(aes(yintercept = threshold_high_centered),
        linetype=2)+
  geom_segment(x = 1961, xend=1990,
               y = 0, yend = 0,
               linewidth=2)+
  scale_fill_hue(l=70, c=60)+
  theme_bw(base_size = 12)+
  ylab("Sommertempratur\navvik fra 1961-1990")+
  xlab("")+
  guides(fill="none")+
  facet_wrap( .~ factor(region, levels = regOrder),
              ncol=3,
              scales = "free_y")
```

Then we can take the mean over the last 5 years and add to a spatial representation.

```{r}
(
i_2022 <- regional_means_long %>%
  filter(year >2017) %>%
  group_by(region) %>%
  summarise(i_2022 = mean(i))
)
```

```{r}
(
sd_2022 <- regional_means_long %>%
  filter(year >2017) %>%
  group_by(region) %>%
  summarise(i_2022 = mean(i))  
)
```
